---
title: "NYC Bike Ridership and Violation Data Analysis"
output: html_document
date: "2023-10-05"
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(knitr)
library(tidyverse)
library(rmarkdown)
library(fpp3)
library(mapview)
library(kableExtra)
```


```{r meta-info-for-future-me, eval=FALSE, include=FALSE}
# This source was essential in helping me post this to github:
# Source: https://mbounthavong.com/blog/2022/7/30/hosting-a-r-markdown-html-file-on-a-github-page
```


```{r load-data, message=FALSE, warning=FALSE, include=FALSE}
df_bicycle_counters_boroughs <- readr::read_csv("/Users/dovwinkleman/Documents/R_Files/NYC_Bike_Summons_and_Violations/Data/Processed/df_bicycle_counters_boroughs.csv")

df_bike_violations <- readr::read_csv("/Users/dovwinkleman/Documents/R_Files/NYC_Bike_Summons_and_Violations/Data/Processed/df_bike_violations_2023-09-16.csv")

df_bike_violations <- df_bike_violations |>
  mutate(violation_code = as.factor(violation_code),
         veh_category = as.factor(veh_category),
         city_nm = as.factor(city_nm),
         rpt_owning_cmd = as.factor(rpt_owning_cmd)
         )

ts_daily_total <- readr::read_csv("/Users/dovwinkleman/Documents/R_Files/NYC_Bike_Summons_and_Violations/Data/Processed/ts_daily_total_2023-10-05.csv")

ts_daily_total<-ts_daily_total |> 
  as_tsibble()
```

# NYC Bike Violations 

## Introduction

Cycling is... While one can find a great deal of information about general cycling in NYC, it is harder to find information specifically about bike tickets.

Throughout this report, I will use the term 'cyclist' to refer to regular bike riders, ebike riders, and escooter riders and 'bicycle' to refer to any of the vehicles that falls under these labels.

## Data

### Datasets

The datasets used were all taken from the NYC Open Data website, except the violation code description data, which was taken from the NY DMV website. The violation data was merged from two different datasets, a 'year to date' dataset spanning Jan 1, 2023 to June 30th, 2023, and a 'historical' dataset dating from Jan 1, 2018 to December 31, 2022. These datasets contained all violations issued by the NYPD, but only the violations labeled 'Bike', 'Ebike', and 'Escooter' were chosen for this report. The bicycle count data was agglomerated per day and per week, and these values were appended to the merged violations dataset. The violation code description data was also merged with the violations dataset, so that a textual description of the violation code could easily be accessed.

### Bike Counters

```{r}
mapview(na.omit(df_bicycle_counters_boroughs), xcol = "longitude", ycol = "latitude", crs = 4269, grid = FALSE)

```
*Bicycle counter locations throughout the 5 boroughs.  Note that two on Amserdam Ave on the west side of Central Park have the exact same latitude and longitude in the .csv file, but in reality one is on Amsterdam ave collecting uptown traffic and the other is an avenue over on Columbus Ave, counting downtown traffic.*

The bicycle count data were taken from 18 separate bike counters, spread throughought the 5 boroughs, as can be seen in the map above.  Although a total of 29 were listed in the .csv file, some of those were either duplicates or not applicable (ie - they only counted pedestrians) and were removed from consideration.  The majority of these counters could be considered ‘Manhattan-centric.’ For instance, there is only one counter in Staten Island by the ferry access to Manhattan, one in a single area in the south Bronx, and only three in Queens, with one of those placed near the Queensboro bridge to Manhattan.  Based on this unequal representation, I chose not to perform any borough specific analyses involving rider count data (however, violation data was not dependent on bike counters), and any generalizations or take-aways from this report should keep these limitations in mind.
The bike counters acted by counting the number of cyclists that crossed them, and recording this aggregation every 15 minutes.  The bike counter data ranged from 2012 to present, but only the data from 2018 to present were used. 

### Removal of NA values

Of  the 126,812 initial violation entries in the specified time frame, 111 total entries were removed, leaving 126,701.  Sixteen rows did not contain violations codes, 92 rows did not contain city name or location information, and 3 rows did not contain location information.


## EDA

### Daily Total Cyclists

```{r hist-daily-total-cyclists, echo=FALSE, message=FALSE, warning=FALSE}
ts_daily_total |>
  ggplot(aes(x = daily_total_cyclists)) +
  geom_histogram(bins = 80)+
  labs(title='Daily Total Cyclists')+
  xlab('Cyclists Per Day')+
  ylab('Count')
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
df_bike_violations |>
  mutate(violation_date = date(violation_date)) |>
  arrange(desc(daily_total_cyclists)) |>
  distinct(violation_date, daily_total_cyclists) |>
  slice(1:10) |>
  kable(caption = "Top 10 Busiest Cycling Days", col.names = linebreak(c('Violation Date', 'Daily Total Cyclists'), align = "l")) |> 
  kable_classic(full_width = F, html_font = "Cambria", position = "left")
```

The amount of total cyclists per day ranged from 1665 to almost 60,000.  The busiest days tended to be in late October/November of 2019/2020, with a few in the middle of summer and a few in September. Further investigation would help uncover why this is, but my guess is that people might be trying to squeeze in one last ride in before winter.


### Daily Totaly Cyclists over Time

```{r}

```


```{r}

```

